Using Model Checking to Analyze the System Behavior of the LHC Production Grid

### 2013-02-04 graphics/ElsevierFGCS_singleColumn.tex
FGCS

The development of DIRAC started in 2002 as a system for production of simulation
data that would serve to verify physics theory, aspects of the LHCb detector design, as
well as to optimize physics algorithms. It gradually evolved into a complete community
grid solution for data and job management, based on a general-purpose framework
that can be reused by other communities besides LHCb. Today, it covers all major
LHCb tasks starting with the raw data transfer from the experiment's detector to
the grid storage, several steps of data processing, up to the final user
analysis. 
It provides a concurrent use of over 60K CPUs and 10M file replicas distributed over
tens of grid sites, provided by the WLCG computing grid \cite{Bird:2011zz}. The community of users has grown to several hundreds, loading the 
LHC grid resources with up to 30K simultaneously running jobs (Fig.~\ref{fig:LHCb-resource-usage}, top) and over 100000 jobs executed per day, during peak processing periods.
About 10PB of data is stored and visible in the LHCb file catalogs, with daily data transfers among storage units reaching 3Gbps (Fig.~\ref{fig:LHCb-resource-usage}, bottom).
\begin{figure*}[tp!]
  \centering
  {\label{fig:Job-throughput}\includegraphics[width=0.7\linewidth]{./graphics/Figure1_1.png}}    
  \vfill
  \vspace{14pt}
{\label{fig:Data-transfers}\includegraphics[width =0.7\linewidth]{./graphics/Figure1_2.png}}
  \caption{LHCb resource usage}
  \label{fig:LHCb-resource-usage}
\end{figure*}
Python was chosen as the implementation language, since it enables
rapid prototyping and development of new features. DIRAC follows the Service
Oriented Architecture (SOA) paradigm, accompanied by a network of lightweight
distributed agents which animate the system. Its main components are depicted in
Fig.~\ref{fig:DIRAC-Arch}.

\begin{figure}[tp]
\includegraphics[width=0.5\textwidth] {./graphics/Figure1.png}
\centering
\caption{DIRAC Architecture overview}
\label{fig:DIRAC-Arch}
\end{figure}

The \textit{services} are passive components that react to requests from their clients,
possibly soliciting other services in order to fulfill the requests. They run as
permanent processes deployed on a number of high-availability hosts (VO-boxes)
at CERN, and store the dynamic system state information in database
repositories. The user interfaces, agents or running jobs can act as clients
placing the requests to DIRAC's services.

@@

\textit{Resources} are software abstractions of the underlying heterogeneous grid
computing and storage entities allocated to LHCb, providing a uniform interface
for access. The physical resources are controlled by the site managers and made
available through middleware services such as gLite \cite{Laure06programmingthe}.

@@

The DIRAC hardware configuration of LHCb is still by far the most powerful one: it consists of about 10 servers (VO-boxes) located at CERN, hosting the central
DIRAC services, including 6 MySQL database servers. Small fraction of the servers and databases are used for testing purposes, and are connected
to the same pool of WLCG computing and storage resources. Each of the six LHCb Tier-1 computing centers accross Europe hosts an additional server as part of this cluster for redundancy,
to ensure high service availability at any moment. Several other experiments are currently using DIRAC on an every day basis, 
with new user groups evaluating its functionality and performance before adopting it in production.

@@

The driving force of DIRAC is the Workload Management System (WMS). Taking into
account the heterogeneous, dynamic, and high-latency nature of the distributed computing environment, a
\textit{Pilot Job paradigm}, illustrated in Fig.~\ref{fig:DIRAC-WMS}, 
\begin{figure}[bp]
\includegraphics[width=0.7\textwidth]{./graphics/Figure2.png}
\centering
\caption{DIRAC Workload Management System \cite{DIRAC_pilot_WMS}}
\label{fig:DIRAC-WMS}
\end{figure}
was chosen as an efficient way to
implement a pull scheduling mechanism.  Pilot jobs are simply resource
reservation processes without an actual payload defined a priori. They are
submitted to the grid worker nodes with the aim of checking the sanity of the
operational environment just before pulling and executing the real payload. This
hides the fragility of the underlying resources and increases the job success
rate, from the perspective of end users. Centralized \textit{Task Queues} contain all the
pending jobs, organized in groups according to their requirements. This enables
efficient application of job priorities, enforcing fair-share policies and
quotas, which are decided within the LHCb community, without the need of any
effort from the grid sites.  Originally, the DIRAC WMS submitted pilot jobs using 
the native gLite middleware tools, but this method was shown to be suboptimal. The gLite resource brokers
are centralized components, and as such they do not have the capacity and flexibility to cope with the load of 
a community such as LHCb. Using multiple resource brokers, rather than a single one, was not the solution:
all brokers use the same site state information and would always submit jobs to the most suitable one globally, causing overloads to
certain sites while underloading others very often. In addition, the gLite information system was not reactive enough to propagate
changes in site status information to the brokers. With the wide deployment of the CREAM CE service \cite{Aiftimiei2010654}, DIRAC's WMS switched
to submitting pilot jobs directly to sites. By interrogating the Computing Element status directly, the system reactiveness
improved dramatically, making sites compete with each other for jobs and have a faster job turnaround.

The basic flowchart describing the evolution of a job's states is depicted in
Fig.~\ref{fig:DIRAC-job-state-machine}. After submission through the \textit{Job Manager} (``Received''), the complete
job description is placed in the DIRAC job repository (the Job DB). Before jobs
become eligible for execution, a chain of optimizer agents checks and prioritizes
them in queues, utilizing the parameters information from the Job DB.
\begin{figure}[tp]
\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure3.png}
\centering
\caption{Job state machine \cite{ProductionShifterGuide}}
\label{fig:DIRAC-job-state-machine}
\end{figure} 
If the requested data resides on tape storage, the \textit{Job
Scheduling Agent} will pass the control to a specialized \textit{Stager} service (part of
the SMS explained in the next section), before placing the job in a Task Queue
(``Waiting''). Based on the complete list of pending payloads, a specialized \textit{Task
Queue Director} agent submits pilots to the computing resources via the available job
submission middleware (e.g., gLite WMS). After
a \textit{Matcher} service pulls the most suitable payload for a pilot (``Matched''), a \textit{Job
Wrapper} object is created on the worker node, responsible for retrieving the
input sandbox, performing software availability checks, executing the actual
payload on the worker node (``Running''), and finally uploading any output data
necessary (``Done'' or ``Failed''). The wrapper can catch any failure exit state
reported by the running physics applications. At the same time, a \textit{Watchdog}
process is instantiated to monitor the behavior of the Job Wrapper and send
heartbeat signals to the monitoring service. It can also take actions in case
resources are soon to be exhausted, the payload stalls, or a management command
for killing the payload is received.

@@

A simplified view of the system is shown in  Fig.~\ref{fig:DIRAC-SMS}. The workflow is initiated
with the \textit{Job Scheduling Agent} detecting that a job is assigned to process files
only available on tape storage. It sends a request for staging (i.e., creating a
cached replica) to the \textit{Storage Manager Handler} service with the
list of files and a callback method to be invoked when the request has been
processed. The Storage Management DB is immediately populated with records which
are processed by a sequence of agents in an organized fashion. The relevant
tables in the SMS DB are the \textit{Tasks} and \textit{CacheReplicas}, whose
entities maintain a state observed and updated by these agents. \textit{Tasks} maintain
general information about every job requesting a service from the SMS. The
details about every file (i.e., the Storage Element where it resides, the
size, checksum, number of tasks that requested it), are kept in the \textit{CacheReplicas}
table. Other auxiliary tables maintain the relationship between these entities.
\begin{figure}[tp]
\begin{center}
 \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure4.png}
\end{center}
\caption{DIRAC Storage Management System}
\label{fig:DIRAC-SMS}
\end{figure}
The processing begins with the \textit{Request Preparation Agent}. It selects all the
``New'' replica entries, checks whether they are registered in a Logical
File Catalog (LFC), and retrieves their metadata. In case of problematic catalog
entries, it can update the state of the \textit{CacheReplicas} and the related \textit{Tasks}
entries to ``Failed''. Non-problematic files are updated to a
``Waiting'' state. The \textit{Stage Request Agent} is responsible for placing the actual staging
requests for all ``Waiting'' entries, via dedicated storage middleware that
communicates with the tape backends. These requests are grouped by Storage Element (SE) prior to
submission, and carry information about the requested (pin) lifetime of the
replicas to be cached. If certain pathologies are discovered (i.e., lost,
unavailable, or zero-sized files on tape), it can update the corresponding
entries to ``Failed'' in a similar manner. Otherwise, they are promoted to
``StageSubmitted''. The agent responsible for monitoring the status of submitted
requests is the \textit{Stage Monitor Agent}. It achieves this by interrogating the
storage middleware to see if the ``StageSubmitted'' files are successfully
replicated on disk cache. In case of success, the \textit{CacheReplicas} and their
corresponding \textit{Tasks} entries are updated to ``Staged''. Various circumstances of
tape or middleware misbehavior can also fail the staging requests. The last one
in the chain is the \textit{Request Finalization Agent}. The \textit{Tasks} which are in their
final states (``Staged'' or ``Failed'') are cleared from the database, and callbacks
are performed to the WMS, which effectively wakes up the corresponding jobs. If
there are no more associated \textit{Tasks} for particular replicas, the respective
\textit{CacheReplicas} entries are also removed.

@@

Apart from describing processes, mCRL2 has a data language that
facilitates describing realistic systems where data influences the
system behavior.  The data language has built-in support for standard
data types such as Booleans (\begin{math}Bool\end{math}), and integer
(\begin{math}Int\end{math}), natural (\begin{math}Nat\end{math})
and positive (\begin{math}Pos\end{math}) numbers, and infinite lists
(\begin{math}List(\_)\end{math}), sets (\begin{math}Set(\_)\end{math})
and bags (\begin{math}Bag(\_)\end{math}) over arbitrary data types. In
addition, users have various ways in which they can define their own
data types and operations on these. For instance, one can define one's
own enumerated data types and reason about these.

@@

High-level design documents are a good starting point, but insufficient for 
building a sound model.
In absence of more detailed up-to-date behavioral models, 
we based our models on the source code and discussions with developers.
A popular abstraction technique
for identifying code subsets that potentially affect particular 
variables of interest (slicing criteria) is program slicing \cite{Hatcliff99slicingsoftware}. 
It reduces the behavior of a program
by removing control statements and data structures deemed irrelevant for 
the criteria. In the context of building a formal model, the slicing criteria 
depends on the focus of the final analysis, i.e., the intended properties to be checked.
The result of such an approach is typically an over-approximation of the original program behavior.
If the abstract model is checked and found to be in conformance with the property,
one can be sure that the behavior of the original program satisfies this property too. 
In other words, the obtained abstraction is sound.

@@

    # Update the states of the files in the database
    if terminalReplicaIDs:
      gLogger.info('%s replicas are terminally failed.' % len( terminalReplicaIDs ) )
      !\colorbox{light-gray}{res = self.stagerClient.}!
	    !\colorbox{light-gray}{updateReplicaFailure(terminalReplicaIDs)}!
    if replicaMetadata:
      gLogger.info('%s replica metadata to be updated.' % len( replicaMetadata ) )
      # Sets the Status='Waiting' of CacheReplicas records 
      #that are OK with catalogue checks
      !\colorbox{light-gray}{res = self.stagerClient.}!
	    !\colorbox{light-gray}{updateReplicaInformation( replicaMetadata )}!
    return S_OK()   
\end{lstlisting}
The logging statements, although critical for operational matters, will
not affect the entries' states, and can be translated to \textit{silent steps} in mCRL2.
Furthermore, instead of tracing back and modeling all variables on which 
the two final lists depend, we can use nondeterminism. 
It is not known upfront which branch execution will follow
for a particular replica, as it depends on external behavior (i.e., the interactions of the system with its environment).
By stubbing out the communication with the LFC and
most of the local variable manipulations 
that follow, and replacing them with a \textit{nondeterministic choice} between the two
ultimate state updates, we can include both possibilities in the model behavior,
and still preserve correctness. Of course, depending on the context, some 
variable values cannot be ignored, in which case determinism can be added
using an \textit{if-then-else} mCRL2 statement. All relevant selection and update 
statements are translated into \textit{actions parametrized with data}.

\subsubsection{Data Abstractions}
The CacheReplicas entity contains more information besides the status of a file.
Every database entry has a unique identifier, descriptive data such as 
the storage where it resides, its full path, checksum, timestamps, etc., as can be seen
from the table description in Fig.~\ref{fig:CacheReplicasTable}.
\begin{figure}[bp!]
\begin{center}
 \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure5.png}
\end{center}
\caption{CacheReplicas table description}
\label{fig:CacheReplicasTable}
\end{figure}
Model checking can only be performed on closed models, where the domains
of all variables are finite. Since we are only interested in state transitions in terms of 
changes to the \textit{Status} field,
we can collapse most of this descriptive data, and represent this entity as a 
\textit{user-defined sort} (type) in mCRL2:
\begin{displaymath}
\begin{aligned}
sort\ CacheReplicas = \textbf{struct}\ Start\ | 
			  New\ |  \\
		    Waiting\ | 
          StageSubmitted\ | \\
		      Staged\ | 
		      Failed\ | 
		  Deleted\ ;
\end{aligned}
\end{displaymath}
This defines an enumerated data type with all possible states.
The Tasks entity is modeled in the same manner. Lists of these sorts can be easily modeled 
in mCRL2 as \begin{math}List(Tasks) \end{math} and 
\begin{math}List(CacheReplicas) \end{math}.
To define the many-to-many relationship between Tasks and CacheReplicas,
we join these data elements in a tuple:
%\begin{lstlisting}[tabsize=10,basicstyle=\ttfamily\fontsize{8}{7}\selectfont]
\begin{displaymath}sort \ Tuple = \textbf{struct}\ p(t:Nat,r:Nat,link:Bool)
\end{displaymath}
%\end{lstlisting}
The first two elements are the list positions of the Tasks and CacheReplicas
entries, while the last one indicates whether a relation between
them exists at a given moment of the system execution.

@@

$ \Gamma_{\{\_RPAgent\_selectCacheReplicas|
   RPAgent\_selectCacheReplicas\_} $
  $  _{\rightarrow RPAgent\_selectCacheReplicas, } $
  $ _{\_RPAgent\_prepareNewReplicas|RPAgent\_prepareNewReplicas\_} $
  $  _{\rightarrow RPAgent\_prepareNewReplicas,\dots\\, } $
 $ _{\_RFAgent\_removeReplicas|RFAgent\_removeReplicas\_} $
  $  _{\rightarrow RFAgent\_removeReplicas\}} $
$ \\ $

@@

  \subfloat[Logging info of a DIRAC job]{\label{fig:Job-Logging-Info}\includegraphics[width=0.4566\linewidth]{./graphics/Figure6a.png}}    

@@

  \subfloat[XSim simulator trace]{\label{fig:simulator}\includegraphics[width =0.5\linewidth]{./graphics/Figure6b.png}}

@@

\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure7_new.png}

@@

on the CacheReplicas memory process, which resulted in only 6 interesting states for this process alone.

The clustering
allowed us to see precisely which chain of actions by the concurrent agents
advanced the state of the CacheReplicas entity. In fact, in the modeling phase we discovered the first problematic
SMS behavior, blocking the progress of staging
tasks (and as a result the progress of jobs). The origin of the deadlock had been a
trivial human logic flaw introduced in coding. At a particular circumstance,
when a ``New'' task enters the system \textbf{and} all associated replicas
are already ``Staged'' (possibly by other jobs requiring the same input files),
its status would immediately be promoted to ``Done''.
Such tasks would never be picked up by SMS and called-back to the appropriate job, 
since the agent responsible for this final step would only look for ``Staged'' tasks.
A proper state machine implementation
instead of hardcoding states in SQL queries can prevent such bugs.
\begin{figure}[bp!]
\vspace{-10 pt}
\includegraphics[width=0.57\linewidth,keepaspectratio=true]{./graphics/Figure8.png}
\centering
\caption{State-space visualisation of the SMS with ltsview}
\label{fig:LTSviewVisualisation}
\end{figure}%123478

@@

\begin{figure}[tp]
\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure9.png}
%was PropertyViolation1.png
\centering
\caption{Violation of requirements 1 (top) and 2 (bottom)}
\label{fig:PropertyViolation}
\end{figure}
Model checking was carried out on a 64bit Intel Core\texttrademark2 Duo (1.6GHz) machine with 2GB RAM. 
Both requirements 1 and 2 were found to be violated, as can be seen from the traces (Fig.~\ref{fig:PropertyViolation}).
The race conditions manifest themselves in a few subtle interactions between the SMS storage and the agents.
In both cases, at step 4 the \textit{Stage Request Agent} issues prestage requests and 
as a result the CacheReplicas entries (second element of the State column) are updated to ``StageSubmitted''. Between the moment
that this agent has selected the corresponding Task to update to the same state,
to the point where the update is done (last step in both traces),
other agents may have monitored these replica records and updated
them to a new state, along with the associated Task. In practice, the manifestation
of such race conditions depends on the speed of the agents propagating the
state changes between the selection and update done by the \textit{Stage Request Agent}.
Such cases are nevertheless encountered in reality, when this agent
has large lists of records to process. This results in a deadlock situation, as 
the Task will have no further state updates made by other agents, since
from their perspective the state propagation is finished.

>>>>

The development of DIRAC started in 2002 as a system for production of simulation
data that would serve to verify physics theory, aspects of the LHCb detector design, as
well as to optimize physics algorithms. It gradually evolved into a complete community
grid solution for data and job management, based on a general-purpose framework
that can be reused by other communities besides LHCb. Today, it covers all major
LHCb tasks starting with the raw data transfer from the experiment's detector to
the grid storage, several steps of data processing, up to the final user
analysis. 
It provides a concurrent use of over 60K CPUs and 10M file replicas distributed over
tens of grid sites. The community of users has grown to several hundreds, loading the 
LHC grid resources with up to 50K simultaneously running jobs during peak processing periods.
Python was chosen as the implementation language, since it enables
rapid prototyping and development of new features. DIRAC follows the Service
Oriented Architecture (SOA) paradigm, accompanied by a network of lightweight
distributed agents which animate the system. Its main components are depicted in
Fig.~\ref{fig:DIRAC-Arch}.

\begin{figure}[tp]
\includegraphics[width=0.5\textwidth] {./Figure1.png}
\centering
\caption{DIRAC Architecture overview}
\label{fig:DIRAC-Arch}
\end{figure}
The \textit{services} are passive components that react to requests from their clients,
possibly soliciting other services in order to fulfill the requests. They run as
permanent processes deployed on a number of high-availability hosts (VO-boxes)
at CERN, and store the dynamic system state information in database
repositories. The user interfaces, agents or running jobs can act as clients
placing the requests to DIRAC's services.

@@

\textit{Resources} are software abstractions of the underlying heterogeneous grid
computing and storage entities allocated to LHCb, providing a uniform interface
for access. The physical resources are controlled by the site managers and made
available through middleware services such as gLite \cite{gLite}.

@@

@@

The driving force of DIRAC is the Workload Management System (WMS). Taking into
account the heterogeneous, dynamic, and high-latency nature of the distributed computing environment, a
\textit{Pilot Job paradigm}, illustrated in Fig.~\ref{fig:DIRAC-WMS}, 
\begin{figure}[bp]
\includegraphics[width=0.7\textwidth]{./Figure2.png}
\centering
\caption{DIRAC Workload Management System \cite{DIRAC_pilot_WMS}}
\label{fig:DIRAC-WMS}
\end{figure}
was chosen as an efficient way to
implement a pull scheduling mechanism.  Pilot jobs are simply resource
reservation processes without an actual payload defined a priori. They are
submitted to the grid worker nodes with the aim of checking the sanity of the
operational environment just before pulling and executing the real payload. This
hides the fragility of the underlying resources and increases the job success
rate, from the perspective of end users. Centralized \textit{Task Queues} contain all the
pending jobs, organized in groups according to their requirements. This enables
efficient application of job priorities, enforcing fair-share policies and
quotas, which are decided within the LHCb community, without the need of any
effort from the grid sites.

The basic flowchart describing the evolution of a job's states is depicted in
Fig.~\ref{fig:DIRAC-job-state-machine}. After submission through the \textit{Job Manager} (``Received''), the complete
job description is placed in the DIRAC job repository (the Job DB). Before jobs
become eligible for execution, a chain of optimizer agents checks and prioritizes
them in queues, utilizing the parameters information from the Job DB.
\begin{figure}[tp]
\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./Figure3.png}
\centering
\caption{Job state machine \cite{ProductionShifterGuide}}
\label{fig:DIRAC-job-state-machine}
\end{figure} 
If the requested data resides on tape storage, the \textit{Job
Scheduling Agent} will pass the control to a specialized \textit{Stager} service (part of
the SMS explained in the next section), before placing the job in a Task Queue
(``Waiting''). Based on the complete list of pending payloads, a specialized \textit{Task
Queue Director} agent submits pilots to the computing resources via the available job
submission middleware (e.g., gLite WMS). After
a \textit{Matcher} service pulls the most suitable payload for a pilot (``Matched''), a \textit{Job
Wrapper} object is created on the worker node, responsible for retrieving the
input sandbox, performing software availability checks, executing the actual
payload on the worker node (``Running''), and finally uploading any output data
necessary (``Done'' or ``Failed''). The wrapper can catch any failure exit state
reported by the running physics applications. At the same time, a \textit{Watchdog}
process is instantiated to monitor the behavior of the Job Wrapper and send
heartbeat signals to the monitoring service. It can also take actions in case
resources are soon to be exhausted, the payload stalls, or a management command
for killing the payload is received.

@@

A simplified view of the system is shown in  Fig.~\ref{fig:DIRAC-SMS}. The workflow is initiated
with the \textit{Job Scheduling Agent} detecting that a job is assigned to process files
only available on tape storage. It sends a request for staging (i.e., creating a
cached replica) to the \textit{Storage Manager Handler} service with the
list of files and a callback method to be invoked when the request has been
processed. The Storage Management DB is immediately populated with records which
are processed by a sequence of agents in an organized fashion. The relevant
tables in the SMS DB are the \textit{Tasks} and \textit{CacheReplicas}, whose
entities maintain a state observed and updated by these agents. \textit{Tasks} maintain
general information about every job requesting a service from the SMS. The
details about every file (i.e., the Storage Element where it resides, the
size, checksum, number of tasks that requested it), are kept in the \textit{CacheReplicas}
table. Other auxiliary tables maintain the relationship between these entities.
\begin{figure}[tp]
\begin{center}
 \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./Figure4.png}
\end{center}
\caption{DIRAC Storage Management System}
\label{fig:DIRAC-SMS}
\end{figure}
The processing begins with the \textit{Request Preparation Agent}. It selects all the
``New'' replica entries, checks whether they are registered in a Logical
File Catalog (LFC), and retrieves their metadata. In case of problematic catalog
entries, it can update the state of the \textit{CacheReplicas} and the related \textit{Tasks}
entries to ``Failed''. Non-problematic files are updated to a
``Waiting'' state. The \textit{Stage Request Agent} is responsible for placing the actual staging
requests for all ``Waiting'' entries, via dedicated storage middleware that
communicates with the tape backends. These requests are grouped by Storage Element (SE) prior to
submission, and carry information about the requested (pin) lifetime of the
replicas to be cached. If certain pathologies are discovered (i.e., lost,
unavailable, or zero-sized files on tape), it can update the corresponding
entries to ``Failed'' in a similar manner. Otherwise, they are promoted to
``StageSubmitted''. The agent responsible for monitoring the status of submitted
requests is the \textit{Stage Monitor Agent}. It achieves this by interrogating the
storage middleware to see if the ``StageSubmitted'' files are successfully
replicated on disk cache. In case of success, the \textit{CacheReplicas} and their
corresponding \textit{Tasks} entries are updated to ``Staged''. Various circumstances of
tape or middleware misbehavior can also fail the staging requests. The last one
in the chain is the \textit{Request Finalization Agent}. The \textit{Tasks} which are in their
final states (``Staged'' or ``Failed'') are cleared from the database, and callbacks
are performed to the WMS, which effectively wakes up the corresponding jobs. If
there are no more associated \textit{Tasks} for particular replicas, the respective
\textit{CacheReplicas} entries are also removed.

@@

Apart from describing processes, mCRL2 has a data language that
facilitate describing realistic systems where data influences the
system behavior.  The data language has built-in support for standard
data types such as Booleans (\begin{math}Bool\end{math}), and integer
(\begin{math}Int\end{math}), natural (\begin{math}Nat\end{math})
and positive (\begin{math}Pos\end{math}) numbers, and infinite lists
(\begin{math}List(\_)\end{math}), sets (\begin{math}Set(\_)\end{math})
and bags (\begin{math}Bag(\_)\end{math}) over arbitrary data types. In
addition, users have various ways in which they can define their own
data types and operations on these. For instance, one can define one's
own enumerated data types and reason about these.

@@

High-level design documents are a good starting point, but insufficient for 
building a sound model.
In absence of more detailed up-to-date behavioral models, 
we based our models on the source code and discussions with developers.
A popular abstraction technique
for identifying code subsets that potentially affect particular 
variables of interest (slicing criteria) is program slicing \cite{Hatcliff99slicingsoftware}. 
It reduces the behavior of a program
by removing control statements and data structures deemed irrelevant for 
the criteria. In the context of building a formal model, the slicing criteria 
depends on the focus of the final analysis, i.e. the intended properties to be checked.
The result of such an approach is typically an over-approximation of the original program behavior.
If the abstract model is checked and found to be in conformance with the property,
one can be sure that the behavior of the original program satisfies this property too. 
In other words, the obtained abstraction is sound.

@@

    # Update the states of the files in the database
    if terminalReplicaIDs:
      gLogger.info('%s replicas are terminally failed.' % len( terminalReplicaIDs ) )
      !\colorbox{light-gray}{res = self.stagerClient.}!
	    !\colorbox{light-gray}{updateReplicaFailure(terminalReplicaIDs)}!
    if replicaMetadata:
      gLogger.info('%s replica metadata to be updated.' % len( replicaMetadata ) )
      # Sets the Status='Waiting' of CacheReplicas records 
      #that are OK with catalogue checks
      !\colorbox{light-gray}{res = self.stagerClient.}!
	    !\colorbox{light-gray}{updateReplicaInformation( replicaMetadata )}!
    return S_OK()   
\end{lstlisting}
The logging statements, although critical for operational matters, will
not affect the entries' states, and can be translated to \textit{silent steps} in mCRL2.
Furthermore, instead of tracing back and modeling all variables on which 
the two final lists depend, we can use nondeterminism. 
It is not known upfront which branch execution will follow
for a particular replica, as it depends on external behavior (i.e. the interactions of the system with its environment).
By stubbing out the communication with the LFC and
most of the local variable manipulations 
that follow, and replacing them with a \textit{nondeterministic choice} between the two
ultimate state updates, we can include both possibilities in the model behavior,
and still preserve correctness. Of course, depending on the context, some 
variable values cannot be ignored, in which case determinism can be added
using an \textit{if-then-else} mCRL2 statement. All relevant selection and update 
statements are translated into \textit{actions parametrized with data}.

\subsubsection{Data Abstractions}
The CacheReplicas entity contains more information besides the status of a file.
Every database entry has a unique identifier, descriptive data such as 
the storage where it resides, its full path, checksum, timestamps, etc., as can be seen
from the table description in Fig.~\ref{fig:CacheReplicasTable}.
\begin{figure}[bp!]
\begin{center}
 \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./Figure5.png}
\end{center}
\caption{CacheReplicas table description}
\label{fig:CacheReplicasTable}
\end{figure}
Model checking can only be performed on closed models, where the domains
of all variables are finite. Since we are only interested in state transitions in terms of 
changes to the \textit{Status} field,
we can collapse most of this descriptive data, and represent this entity as a 
\textit{user-defined sort} (type) in mCRL2:
\begin{displaymath}
\begin{aligned}
sort\ CacheReplicas = \textbf{struct}\ Start\ | 
			  New\ |  \\
		    Waiting\ | 
          StageSubmitted\ | \\
		      Staged\ | 
		      Failed\ | 
		  Deleted\ ;
\end{aligned}
\end{displaymath}
This defines an enumerated data type with all possible states.
The Tasks entity is modeled in the same manner. Lists of these sorts can be easily modeled 
in mCRL2 as \begin{math}List(Tasks) \end{math} and 
\begin{math}List(CacheReplicas) \end{math}.
To define the many-to-many relationship between Tasks and CacheReplicas,
we join these data elements in a tuple:
%\begin{lstlisting}[tabsize=10,basicstyle=\ttfamily\fontsize{8}{7}\selectfont]
\begin{displaymath}sort \ Tuple = \textbf{struct}\ p(t:Nat,r:Nat,link:Bool)
\end{displaymath}
%\end{lstlisting}
The first two elements are the list positions of the Tasks and CacheReplicas
entries, while the last one indicates whether a relation between
them exists at a given moment of the system execution.

@@

$ \Gamma_{\{\_RPAgent\_selectCacheReplicas|
   RPAgent\_selectCacheReplicas\_} $
  $  _{\rightarrow RPAgent\_selectCacheReplicas, } $
$\\$
  $ _{\_RPAgent\_prepareNewReplicas|RPAgent\_prepareNewReplicas\_} $
  $  _{\rightarrow RPAgent\_prepareNewReplicas,\dots\\, } $
$\\$
 $ _{\_RFAgent\_removeReplicas|RFAgent\_removeReplicas\_} $
  $  _{\rightarrow RFAgent\_removeReplicas\}} $
$ \\ $

@@

  \subfloat[Logging info of a DIRAC job]{\label{fig:Job-Logging-Info}\includegraphics[width=0.4566\linewidth]{./Figure6a.png}}    

@@

  \subfloat[XSim simulator trace]{\label{fig:simulator}\includegraphics[width =0.5\linewidth]{./Figure6b.png}}

@@

\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./Figure7.png}

@@

on the CacheReplicas memory process, which resulted in only 7 interesting states for this process alone.

The clustering
allowed us to see precisely which chain of actions by the concurrent agents
advanced the state of the CacheReplicas entity. In fact, in the modeling phase we discovered the first problematic
SMS behavior, blocking the progress of staging
tasks (and as a result the progress of jobs). The origin of the deadlock had been a
trivial human logic flaw introduced in coding. At a particular circumstance,
when a ``New'' task enters the system \textbf{and} all associated replicas
are already ``Staged'' (possibly by other jobs requiring the same input files),
its status would immediately be promoted to ``Done''.
Such tasks would never be picked up by SMS and called-back to the appropriate job, 
since the agent responsible for this final step would only look for ``Staged'' tasks.
A proper state machine implementation
instead of hardcoding states in SQL queries can prevent such bugs.
\begin{figure}[bp!]
\vspace{-10 pt}
\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./Figure8.png}
\centering
\caption{State-space visualisation of the SMS with ltsview}
\label{fig:LTSviewVisualisation}
\end{figure}%123478

@@

\begin{figure}[tp]
\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./Figure9.png}
%was PropertyViolation1.png
\centering
\caption{Violation of requirements 1 (top) and 2 (bottom)}
\label{fig:PropertyViolation}
\end{figure}
Model checking was carried out on a 64bit Intel Core\texttrademark2 Duo (1.6GHz) machine with 2GB RAM. 
Both requirements 1 and 2 were found to be violated, as can be seen from the traces (Fig.~\ref{fig:PropertyViolation}).
The race conditions manifest themselves in a few subtle interactions between the SMS storage and the agents.
In both cases, at step 4 the \textit{Stage Request Agent} issues prestage requests and 
as a result the CacheReplicas entries (second element of the State column) are updated to ``StageSubmitted''. Between the moment
that this agent has selected the corresponding Task to update to the same state,
to the point where the update is done (last step in both traces),
other agents may have monitored these replica records and updated
them to a new state, along with the associated Task. In practice, the manifestation
of such race conditions depends on the speed of the agents propagating the
state changes between the selection and update done by the \textit{Stage Request Agent}.
Such cases are nevertheless encountered in reality, when this agent
has large lists of records to process. This results in a deadlock situation, as 
the Task will have no further state updates made by other agents, since
from their perspective the state propagation is finished.
